Experience in Modeling a Microcontroller Instruction Set Using B

### 2012-01-22 trunk/papers/ArticleZ80/source/paper.tex
Pequenos ajustes na conclusão e outras correções sugeridas pelo professor de inglês.

In general, the manuals of microcontrollers can be more suited for \marginpar{DD: o título da sub-seção não condiz com seu conteúdo. Não cita nenhum problema na formalização. Não vejo muito bem o que ele traz de novo que já não foi dito anteriormente.}
developers in assembly language. Many times, each instruction of
microcontrollers is shown in its official manual using textual
description, math description, examples, encoding, number of cycles
and other information. Sometimes, the instruction descriptions have
an own notation, some informations are organized in different sections
of document and the textual description are not standardized. 
The manuals also have semi-formal and informal descriptions, these descriptions
may leave errors, inconsistency and ambiguity.  
For example in official manual of the microcontroller Z80 \cite{Z80_manual}, if the users \marginpar{VM: Adicionei um exemplo real e uma referência  que cita vários problemas identificados no Manual do Z80.} 
want to know the action of an instruction and its effects on the flag register then he needs
to read an informal textual description, that  some times can be placed in different pages. 
Furthermore, the official manual of the microcontroller Z80 has also several
problems that were identified over time~\cite{UndocumentedZ80}.

@@

This section presents the verification of the aforementioned modules
using the AtelierB provers.  By default, AtelierB~\cite{atelierB} does
not generate proof obligations of the type of functions defined in the
clause $PROPERTIES$ of a machine. For example, the definition $inc \in
\{0,1\} \fun \{0,1\} \land inc = \lambda(x).( x \in \{0,1\} | x + 1)$,
does not create a proof obligation to check the consistency of the
definition. However, this definition is inconsistent since $inc(1)=2$
and $ 2 \not\in \{0,1\}$.

@@

%\textbf{We needed to verify the type of these conversion
%  functions}.\marginpar{Não precisa repetir
%  isto.} % There are twenty rules associated this properties
The conversion functions need to satisfy two properties: to be
bijective and equal to the reverse of the dual conversion
function. The three simplest proof obligations were verified
automatically with the AtelierB: (\it uchar\_schar\rm \ $\in$
\textit{UCHAR} $\bij$ \textit{SCHAR}), (\it schar\_uchar\rm \ $\in$
\textit{SCHAR} $\bij$ \textit{UCHAR}) and (\it byte\_bv16\rm \ $=$ \it
bv16\_byte\rm $^{-1}$).
%\begin{enumerate}
%
%\item (\it uchar\_schar\rm \ $\in$ \textit{UCHAR} $\bij$ \textit{SCHAR})
%\item (\it schar\_uchar\rm \ $\in$ \textit{SCHAR} $\bij$ \textit{UCHAR})
%\item (\it byte\_bv16\rm \ $=$ \it bv16\_byte\rm $^{-1}$) 
%
%\end{enumerate}
For the remaining proof obligations, we again had to create new proof
rules. There are three kinds of proof rules, that we illustrate by giving
one example each:

@@

	\bf SETS\\
	\hspace*{0.10in}\it id\_reg\_8 \rm = \rm \{ \it a0\rm, \it f0\rm,\it b0\rm, \it c0\rm,\it d0\rm, \it e0\rm,\it h0\rm, \it l0\rm,\it a\_0\rm, \it f\_0\rm,\it b\_0\rm, \it c\_0\rm,\it d\_0\rm, \it e\_0\rm,\it h\_0\rm, \it l\_0 \rm \};\\
	\hspace*{0.10in}\it id\_reg\_16 \rm = \rm \{ \it BC\rm, \it DE\rm, \it HL\rm, \it SP\rm, \it AF \rm \}

@@

  \item The mode 0 is compatible with 8080 and this mode is selected when \it im \rm = \rm ( \rm 0 $\mapsto$  \rm 0 \rm ). When a 
  non-maskable interrupt happens, the CPU fetches an instruction of one byte
  from an external device, usually an RST instruction, and the CPU executes it.
  The instruction code is received from an external device by data bus and it is represented by integer parameter called $\textit{byte\_bus}$ .

@@

A case study using the first version of Z80 model was also created and verified
since the first abstract B model up to B assembly model. This case study is an
object of the pilot project developed to analyse a petroleum production test
system in each oil field. Its modeling was developed in according to \cite{LAUT_SERGIO}
and its B assembly model can be executed manually in ProB. The modeling of
instruction set and the B assembly model allow to simulate the execution of microcontroller
and analyse several important properties.

@@

There are in the literature several similar approaches to model hardware and instruction set using B.
The paper \cite{springerlink:Yuan2011} reports a method to specify, design and construct sound and complete instruction set model by stepwise refinement and formal proof using the formal method Event-B. It also discuss about desirable properties of instruction set model. However, our work \cite{Valerio_SBMF09} and \cite{Subotic2010} use the B method, that seems more appropriated to software development, because it has implementable language defined, 
called B0, and tools to convert the models to a programming language.
In \cite{Marc20113} is present an related experience on using B in the design of secure micro-controllers. It tries to show the feasibility of such a technique for high confidence trustful devices. The work \cite{Subotic2010} describes a similar approach with MIPS CPU architecture and it shows how the CPU is formally specified and implemented using the B method.

@@

% Um pequeno resumo do trabalho
This work has shown an approach to the formal modeling of the instruction set of
microcontroller using the B method. During the construction of this model, some
problems were found in the official reference for Z80
microcontroller~\cite{Z80_manual}. The designer fixed the found problems in
documentation and also developed a case study: a verified embedded software. This
case study was interesting to analyse the technique of formal verification up to
assembly level.

% TODO: Adaptar para apenas fundamentar melhor o trabalho já desenvolvido. Citar o artigo SBMF e a dissertação. 
% Os meus trabalhos já desenvolvidos/ resultados atuais.
The next works quoted are directly related to the objective this paper.
The first work \cite{DantasSemish2008} shows the developed metodology to verify
software up to assembly level using B. A second work\cite{Dantas_SBMF08}
presents more details about the verification approach and a small example
software verified up to assembly level in three different platforms. 
The last work \cite{Valerio_SBMF09} shows a general view of the first version of Z80
model and techniques used in verification process.
%Link com o parágrafo anterior 
These works created a important step to software verification up to assembly
language, they show the actuals difficulties and suggest improvements in the
B tools.

>>>>

Many times, each instruction of microcontrollers is shown in its official manual
using textual description, math description, examples, encoding, number of cycles
and other information. Sometimes, the instruction descriptions have
an own notation, some informations are organized in different sections
of document and the textual description are not standardized. 
The manuals also have semi-formal and informal descriptions, these descriptions
may leave errors, inconsistency and ambiguity.  
For example in official manual of the microcontroller Z80 \cite{Z80_manual}, if the users \marginpar{VM: Adicionei um exemplo real e uma referência  que cita vários problemas identificados no Manual do Z80.} 
want to know the action of an instruction and its effects on the flag register then he needs
to read an informal textual description, that  sometimes can be placed in different pages. 
Furthermore, the official manual of the microcontroller Z80 has also several
problems that were identified over time~\cite{UndocumentedZ80}.

@@

This section presents the verification of the aforementioned modules
using AtelierB provers.  By default, AtelierB~\cite{atelierB} does
not generate proof obligations of the type of functions defined in the
clause $PROPERTIES$ of a machine. For example, the definition $inc \in
\{0,1\} \fun \{0,1\} \land inc = \lambda(x).( x \in \{0,1\} | x + 1)$,
does not create a proof obligation to check the consistency of the
definition. However, this definition is inconsistent since $inc(1)=2$
and $ 2 \not\in \{0,1\}$.

@@

%\textbf{We needed to verify the type of these conversion
%  functions}.\marginpar{Não precisa repetir
%  isto.} % There are twenty rules associated this properties
The conversion functions need to satisfy two properties: to be
bijective and equal to the reverse of the dual conversion
function. The three simplest proof obligations were verified
automatically with AtelierB: (\it uchar\_schar\rm \ $\in$
\textit{UCHAR} $\bij$ \textit{SCHAR}), (\it schar\_uchar\rm \ $\in$
\textit{SCHAR} $\bij$ \textit{UCHAR}) and (\it byte\_bv16\rm \ $=$ \it
bv16\_byte\rm $^{-1}$).
%\begin{enumerate}
%
%\item (\it uchar\_schar\rm \ $\in$ \textit{UCHAR} $\bij$ \textit{SCHAR})
%\item (\it schar\_uchar\rm \ $\in$ \textit{SCHAR} $\bij$ \textit{UCHAR})
%\item (\it byte\_bv16\rm \ $=$ \it bv16\_byte\rm $^{-1}$) 
%
%\end{enumerate}
For the remaining proof obligations, we again had to create new proof
rules. There are three kinds of proof rules, that we illustrate by giving
one example each:

@@

	\bf SETS\\
\hspace*{0.10in}\it id\_reg\_8 \rm = \rm \{ \it a0\rm, \it f0\rm, \it b0\rm, \it
c0\rm, \it d0\rm, \it e0\rm, \it h0\rm, \it l0\rm, \it a\_0\rm,  \it f\_0\rm,
\it b\_0\rm, \it c\_0\rm, \it d\_0\rm, \it e\_0\rm, \it h\_0\rm, \it l\_0 \rm
\};\\ \hspace*{0.10in}\it id\_reg\_16 \rm = \rm \{  \it AF \rm, \it BC\rm, \it
DE\rm, \it HL\rm, \it SP\rm\}

@@

  \item The mode 0 is compatible with 8080 and this mode is selected when \it im \rm = \rm ( \rm 0 $\mapsto$  \rm 0 \rm ). When a  non-maskable interrupt happens, the CPU fetches an instruction of one byte
  from an external device, usually an RST instruction, and the CPU executes it.
  The instruction code is received from an external device by data bus and it is represented by integer parameter called $\textit{byte\_bus}$.

@@

A case study using the first version of Z80 model was also created and verified
since the first abstract B model up to B assembly model. This case study is an
object from the pilot project developed to analyse a petroleum production test
system in each oil field. Its modeling was developed according to \cite{LAUT_SERGIO}
and its B assembly model can be executed manually in ProB. The modeling of
instruction set and the B assembly model allow to simulate the execution of microcontroller
and analyse several important properties.

@@

There are in the literature several similar approaches to model hardware and instruction set using B.
The paper \cite{springerlink:Yuan2011} reports a method to specify, design and construct sound and complete instruction set model by stepwise refinement and formal proof using Event-B. It also discusses about desirable properties of instruction set model. However, our work \cite{Valerio_SBMF09} and \cite{Subotic2010} use the B method, that seems quite appropriated to software development, because it has 
an implementable language defined, called B0, a model-driven approach \cite{Dantas_SBMF08} to develop software from the functional specification level down to assembly, and tools to convert the models to a programming language. An related experience on using B in the design of secure micro-controllers is present in \cite{Marc20113}. It tries to show the feasibility of such a technique for high confidence trustful devices. The work \cite{Subotic2010} describes a similar approach with MIPS CPU architecture and it shows how the CPU is formally specified and implemented using the B method.

@@

% Um pequeno resumo do trabalho
This work has shown an approach to the formal modeling of the instruction set of
microcontroller using the B method.  During the construction of this model, some
problems were found in the official reference for Z80
microcontroller~\cite{Z80_manual}. We have consulted different sources \cite{Simulator_z80,UndocumentedZ80,Z80_manual}  to avoid problems in modeling and have developed a case study: a verified embedded software.
This case study was interesting to analyse the technique of formal verification up to assembly level.

% TODO: Adaptar para apenas fundamentar melhor o trabalho já desenvolvido. Citar o artigo SBMF e a dissertação. 
% Os meus trabalhos já desenvolvidos/ resultados atuais.
The following works quoted are directly related to the objective this paper.
The first work \cite{DantasSemish2008} shows the developed metodology to verify
software up to assembly level using B. A second work\cite{Dantas_SBMF08}
presents more details about the verification approach and a small 
software that was verified up to assembly level in three different platforms. 
The last work \cite{Valerio_SBMF09} shows a general view of the first version of Z80
model and techniques used in verification process. Basically, the new Z80 model presented here has added support for animation in ProB and the representation of external actions (interruptions).
%Link com o parágrafo anterior 
These works created an important step to software verification up to assembly
language, they show the current difficulties and suggest improvements in the
B tools.
